---
title: "Analyzing Play-by-Play Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analyzing Play-by-Play Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo = FALSE, message = FALSE}
library(tibble)
library(dplyr)
```

> **RETROSHEET DATA NOTICE**
>
> The information used here was obtained free of charge from and is copyrighted by Retrosheet. Interested parties may contact Retrosheet at 20 Sunset Rd., Newark, DE 19711.

Play-by-play events are what make Retrosheet special: no other freely available source describes every play of (nearly) every major league game for over a century. This article shows how to turn that notation into batting statistics, splits, and historical reconstructions with `parse_plays()`.

Because everything downloads from Retrosheet, the code chunks are not evaluated at build time; the tables shown are real outputs from the 2024 and 1954 postseasons.

## From notation to typed columns

`get_plays()` returns Retrosheet's raw event notation (`"S7/G56"`, `"64(1)3/GDP"`). `parse_plays()` interprets that notation and appends typed columns:

```{r, eval = FALSE}
library(retrosheetshow)
library(dplyr)

plays <- get_events(year = 2024, type = "post") |>
  get_plays() |>
  parse_plays()

plays |>
  select(event, event_type, is_hit, hit_value, trajectory, outs_on_play, rbi)
```

```{r, echo = FALSE}
tribble(
  ~event, ~event_type, ~is_hit, ~hit_value, ~trajectory, ~outs_on_play, ~rbi,
  "7/F7S", "generic_out", FALSE, 0L, "fly", 1L, 0L,
  "S7/G56", "single", TRUE, 1L, "ground", 0L, 0L,
  "K", "strikeout", FALSE, 0L, NA, 1L, 0L,
  "HR/F78XD.2-H;1-H", "home_run", TRUE, 4L, "fly", 0L, 3L
) |>
  knitr::kable(caption = "Parsed play columns (selected 2024 postseason plays)")
```

The parser follows the published grammar at <https://www.retrosheet.org/eventfile.htm> and is validated against Chadwick's `cwevent`, the community reference implementation (see the package's test suite). A season of plays parses in a few seconds.

What happened, across the whole 2024 postseason:

```{r, eval = FALSE}
plays |>
  count(event_type, sort = TRUE)
```

```{r, echo = FALSE}
tribble(
  ~event_type, ~n,
  "generic_out", 1459L,
  "strikeout", 739L,
  "no_play", 599L,
  "single", 443L,
  "walk", 309L,
  "double", 107L,
  "home_run", 98L,
  "stolen_base", 57L
) |>
  knitr::kable(caption = "Most common events, 2024 postseason (top 8)")
```

(`no_play` records mark substitutions; filter them out or ignore them, since all their statistic columns are zero.)

## Batting lines from raw plays

The flag columns follow official scoring rules — sacrifices and catcher's interference are plate appearances but not at bats, walks never count as at bats — so slash lines are a single `summarize()`:

```{r, eval = FALSE}
batting <- plays |>
  summarize(
    pa = sum(is_plate_appearance),
    ab = sum(is_at_bat),
    h = sum(is_hit),
    hr = sum(event_type == "home_run"),
    rbi = sum(rbi),
    bb = sum(event_type %in% c("walk", "intentional_walk")),
    hbp = sum(event_type == "hit_by_pitch"),
    tb = sum(hit_value),
    sf = sum(is_sac_fly),
    .by = player_id
  ) |>
  mutate(
    avg = h / ab,
    obp = (h + bb + hbp) / (ab + bb + hbp + sf),
    slg = tb / ab
  )
```

Player IDs become names by joining the biographical database:

```{r, eval = FALSE}
players <- get_player_ids()

batting |>
  filter(pa >= 40) |>
  left_join(
    players |> select(player_id, first_name, last_name),
    by = "player_id"
  ) |>
  arrange(desc(slg)) |>
  select(first_name, last_name, pa, h, hr, rbi, avg, obp, slg)
```

```{r, echo = FALSE}
tribble(
  ~first_name, ~last_name, ~pa, ~h, ~hr, ~rbi, ~avg, ~obp, ~slg,
  "Giancarlo", "Stanton", 62L, 15L, 7L, 16L, 0.273, 0.339, 0.709,
  "Mark", "Vientos", 58L, 18L, 5L, 14L, 0.327, 0.362, 0.636,
  "Juan", "Soto", 64L, 16L, 4L, 9L, 0.327, 0.469, 0.633,
  "Pete", "Alonso", 58L, 12L, 4L, 10L, 0.273, 0.431, 0.568,
  "Mookie", "Betts", 75L, 18L, 4L, 16L, 0.290, 0.387, 0.565
) |>
  knitr::kable(
    digits = 3,
    caption = "2024 postseason slugging leaders (minimum 40 PA)"
  )
```

## Situational splits

Each play carries the count when the event happened, so count-based splits need no extra data. Two strikes changes everything:

```{r, eval = FALSE}
plays |>
  filter(is_plate_appearance, count != "??") |>
  mutate(two_strikes = substr(count, 2, 2) == "2") |>
  summarize(
    pa = n(),
    avg = sum(is_hit) / sum(is_at_bat),
    slg = sum(hit_value) / sum(is_at_bat),
    k_rate = mean(event_type == "strikeout"),
    .by = two_strikes
  )
```

```{r, echo = FALSE}
tribble(
  ~two_strikes, ~pa, ~avg, ~slg, ~k_rate,
  FALSE, 1526L, 0.318, 0.513, 0,
  TRUE, 1740L, 0.156, 0.263, 0.425
) |>
  knitr::kable(
    digits = 3,
    caption = "2024 postseason hitting before vs. with two strikes"
  )
```

Other splits that work directly from the parsed columns: by inning, by home/visiting team (`team` is 0/1), bunts vs. full swings (`is_bunt`), or ball-in-play trajectory (`trajectory`, `hit_location`).

## A century of history: the 1954 World Series

Retrosheet's postseason coverage reaches back to 1903, and the same three lines of code work all the way down. Game 1 of the 1954 World Series contains perhaps the most famous defensive play ever made — Willie Mays' over-the-shoulder catch off Vic Wertz, with the game tied in the eighth:

```{r, eval = FALSE}
plays_1954 <- get_events(year = 1954, type = "post") |>
  get_plays() |>
  parse_plays()

plays_1954 |>
  filter(game_id == "NY1195409290", inning == 8, team == 0, event != "NP") |>
  select(player_id, event, event_type, trajectory, hit_location, outs_on_play)
```

```{r, echo = FALSE}
tribble(
  ~player_id, ~event, ~event_type, ~trajectory, ~hit_location, ~outs_on_play,
  "dobyl101", "W", "walk", NA, NA, 0L,
  "rosea101", "S6.1-2", "single", NA, NA, 0L,
  "wertv101", "8/F8XD.2-3", "generic_out", "fly", "8XD", 1L,
  "mitcd101", "W.1-2", "walk", NA, NA, 0L,
  "poped101", "K/C", "strikeout", NA, NA, 1L,
  "hegaj101", "7", "generic_out", NA, NA, 1L
) |>
  knitr::kable(caption = "Top of the 8th, 1954 World Series game 1")
```

The third row is The Catch: a fly out to center field, extra deep (`8XD`), with the runner from second only reaching third — a ball that leaves any other park. Two innings later the notation records how the game ended:

```{r, eval = FALSE}
plays_1954 |>
  filter(game_id == "NY1195409290", event_type == "home_run") |>
  select(player_id, inning, event, runs_on_play, rbi)
```

```{r, echo = FALSE}
tribble(
  ~player_id, ~inning, ~event, ~runs_on_play, ~rbi,
  "rhodd101", 10L, "HR/9D.2-H;1-H", 3L, 3L
) |>
  knitr::kable(caption = "Dusty Rhodes' pinch-hit walk-off, 3 runs, 3 RBI")
```

Note the era differences: 1950s files rarely record counts, pitch sequences, or batted-ball trajectories, so those columns are `NA` when the source does not have them. Everything that is in the notation — events, outs, runs, RBI, advancement — parses the same in 1954 as in 2024.

## Game state: outs, runners, and score

`parse_plays()` reads one event string at a time, so it cannot know what the bases looked like when the play began. `track_game_state()` can: it replays each game's full record stream — plays, substitutions, and the automatic extra-inning runner — and returns every play with the situation it started from:

```{r, eval = FALSE}
states <- get_events(year = 2024, type = "post") |>
  track_game_state()
```

The most famous at-bat of the 2024 postseason, as the tracker sees it — Freddie Freeman stepping in with the World Series opener on the line:

```{r, eval = FALSE}
states |>
  filter(game_id == "LAN202410250", inning == 10, event_type == "home_run") |>
  select(player_id, event, outs_before,
         runner1_id, runner2_id, runner3_id, bat_score, fld_score)
```

```{r, echo = FALSE}
tribble(
  ~player_id, ~event, ~outs_before, ~runner1_id, ~runner2_id, ~runner3_id, ~bat_score, ~fld_score,
  "freef001", "HR/F9D.3-H;2-H;1-H", 2L, "bettm001", "edmat001", "taylc001", 2L, 3L
) |>
  knitr::kable(
    caption = "Freeman's walk-off grand slam: two outs, bases loaded, down a run"
  )
```

State columns are validated against Chadwick's `cwevent` the same way the parser is: on the golden files and in full-season cross-checks, outs, scores, and all three runner identities agreed on every one of ~283,000 events. Knowing the out count also lets `track_game_state()` apply the one scoring rule the string-level parser cannot — no RBI for a runner from third scoring on an error with two outs — so its `rbi` column is exact.

## Run expectancy

With base-out states attached, the classic RE24 matrix is a dozen lines: for each play, count the runs scored from that point to the end of the half-inning, then average by starting state.

```{r, eval = FALSE}
states <- get_events(year = 2024) |>
  track_game_state()

re24 <- states |>
  mutate(
    bases = paste0(
      if_else(is.na(runner1_id), "-", "1"),
      if_else(is.na(runner2_id), "-", "2"),
      if_else(is.na(runner3_id), "-", "3")
    )
  ) |>
  mutate(
    runs_rest = rev(cumsum(rev(runs_on_play))),
    complete = sum(outs_on_play) == 3,
    .by = c(game_id, inning, team)
  ) |>
  filter(complete) |>
  summarize(re = mean(runs_rest), .by = c(bases, outs_before))

re24 |>
  tidyr::pivot_wider(
    id_cols = bases, names_from = outs_before,
    values_from = re, names_prefix = "outs_"
  ) |>
  arrange(bases)
```

```{r, echo = FALSE}
tribble(
  ~bases, ~outs_0, ~outs_1, ~outs_2,
  "---", 0.48, 0.25, 0.09,
  "--3", 1.36, 0.93, 0.34,
  "-2-", 1.08, 0.65, 0.33,
  "-23", 1.98, 1.37, 0.58,
  "1--", 0.89, 0.52, 0.22,
  "1-3", 1.85, 1.17, 0.48,
  "12-", 1.46, 0.92, 0.43,
  "123", 2.25, 1.61, 0.80
) |>
  knitr::kable(
    digits = 2,
    caption = "Run expectancy by base-out state, 2024 regular season"
  )
```

Freeman's grand slam, in these terms: bases loaded with two outs is worth 0.80 expected runs, and he delivered four. The same state table is the foundation for RE24 batting value, stolen-base break-even rates, and bunting analysis — all now a `summarize()` away.

## What the string-level parser alone cannot know

If you use `parse_plays()` without `track_game_state()`, keep in mind:

* `runner1_dest`, `runner2_dest`, and `runner3_dest` are `NA` when the play does not mention that runner — a runner who holds is indistinguishable from an empty base. `track_game_state()` resolves this with its `runner*_id` columns.
* The `rbi` column defaults to crediting the rule 9.04(a)(3) edge case (runner from third scoring on an error with two outs); `track_game_state()` corrects it. In two full validation seasons this affected 2 plays in about 283,000.

## See Also

- `?parse_plays` and `?track_game_state` for the full column references
- The **Retrosheet Format Reference** article for the notation itself
- The **Getting Started** article for downloading and extracting events

## Attribution

Remember to cite Retrosheet when using this data:

> The information used here was obtained free of charge from and is copyrighted by Retrosheet. Interested parties may contact Retrosheet at 20 Sunset Rd., Newark, DE 19711.
