---
title: "Game Logs, Rosters, and Schedules"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Game Logs, Rosters, and Schedules}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo = FALSE, message = FALSE}
library(tibble)
library(dplyr)
```

This article demonstrates retrosheetshow's game logs, rosters, and schedules features. Because everything downloads from Retrosheet, the code chunks are not evaluated at build time; the tables shown are real outputs from the 2024 season.

## Game Logs

Game logs provide summary statistics for each game (one row per game, 161 columns). They're much faster to download than full play-by-play events.

### Download Game Logs

```{r, eval = FALSE}
library(retrosheetshow)
library(dplyr)

gamelogs_2024 <- get_gamelogs(year = 2024)
gamelogs_2024 |>
  select(date, visiting_team, home_team, visiting_score, home_score,
         park_id, attendance)
```

```{r, echo = FALSE}
tribble(
  ~date, ~visiting_team, ~home_team, ~visiting_score, ~home_score, ~park_id, ~attendance,
  "20240320", "LAN", "SDN", 5L, 2L, "SEO01", 15952L,
  "20240321", "SDN", "LAN", 15L, 11L, "SEO01", 15952L,
  "20240328", "COL", "ARI", 1L, 16L, "PHO01", 48559L,
  "20240328", "WAS", "CIN", 2L, 8L, "CIN09", 43799L
) |>
  knitr::kable(caption = "2024 game logs (selected columns, first 4 games)")
```

Scores, counting statistics, attendance, and game numbers are parsed as integers, so they can be compared and aggregated directly.

### Home Field Advantage

```{r, eval = FALSE}
gamelogs_2024 |>
  summarize(
    total_games = n(),
    home_win_pct = mean(home_score > visiting_score)
  )
```

```{r, echo = FALSE}
tribble(
  ~total_games, ~home_win_pct,
  2429L, 0.522
) |>
  knitr::kable(caption = "Home teams won 52.2% of games in 2024")
```

### Highest Scoring Games

```{r, eval = FALSE}
gamelogs_2024 |>
  mutate(total_runs = visiting_score + home_score) |>
  arrange(desc(total_runs)) |>
  select(date, visiting_team, home_team, visiting_score, home_score, total_runs) |>
  head(5)
```

```{r, echo = FALSE}
tribble(
  ~date, ~visiting_team, ~home_team, ~visiting_score, ~home_score, ~total_runs,
  "20240706", "BAL", "OAK", 8L, 19L, 27L,
  "20240724", "BOS", "COL", 7L, 20L, 27L,
  "20240321", "SDN", "LAN", 15L, 11L, 26L,
  "20240612", "COL", "MIN", 9L, 17L, 26L,
  "20240826", "CHN", "PIT", 18L, 8L, 26L
) |>
  knitr::kable(caption = "Top 5 highest scoring games in 2024")
```

### Top Winning Pitchers

```{r, eval = FALSE}
gamelogs_2024 |>
  count(winning_pitcher_name, sort = TRUE) |>
  head(5)
```

```{r, echo = FALSE}
tribble(
  ~winning_pitcher_name, ~n,
  "Chris Sale", 18L,
  "Tarik Skubal", 18L,
  "Carlos Rodon", 16L,
  "Jose Berrios", 16L,
  "Seth Lugo", 16L
) |>
  knitr::kable(caption = "Pitchers credited with the most wins in 2024")
```

### Field Descriptions

```{r, eval = FALSE}
# See all 161 available fields
fields <- gamelog_fields()

# Find fields related to home runs
fields |>
  filter(grepl("hr", field_name))
```

```{r, echo = FALSE}
tribble(
  ~field_name, ~description,
  "visiting_hr", "Visiting home runs",
  "home_hr", "Home home runs"
) |>
  knitr::kable(caption = "Home run fields in game logs")
```

## Rosters

Rosters are extracted from the event archives (and reuse the same cache), with one row per player per team.

```{r, eval = FALSE}
# All 2024 rosters
rosters_2024 <- get_rosters(year = 2024)

# One team
yankees <- get_rosters(year = 2024, team = "NYA")
yankees |> arrange(last_name) |> head(4)
```

```{r, echo = FALSE}
tribble(
  ~player_id, ~last_name, ~first_name, ~bats, ~throws, ~team, ~position,
  "andrc002", "Andrews", "Clayton", "L", "L", "NYA", "P",
  "beetc001", "Beeter", "Clayton", "R", "R", "NYA", "P",
  "bertj001", "Berti", "Jon", "R", "R", "NYA", "3B",
  "bickp001", "Bickford", "Phil", "R", "R", "NYA", "P"
) |>
  knitr::kable(caption = "2024 Yankees roster (first 4 players)")
```

### Multi-Year Tracking

```{r, eval = FALSE}
# Track roster changes over time
multi_year_rosters <- get_rosters(year = 2022:2024, team = "NYA")

# Find players on the roster in multiple years
multi_year_rosters |>
  count(player_id, last_name, first_name, sort = TRUE) |>
  filter(n > 1)
```

## Schedules

Schedules contain the planned games for a season, including postponement information.

```{r, eval = FALSE}
schedule_2024 <- get_schedules(year = 2024)
schedule_2024 |>
  select(date, day_of_week, visiting_team, home_team, day_night, postponed)
```

```{r, echo = FALSE}
tribble(
  ~date, ~day_of_week, ~visiting_team, ~home_team, ~day_night, ~postponed,
  "20240320", "Wednesday", "LAN", "SDN", "n", NA,
  "20240321", "Thursday", "SDN", "LAN", "n", NA,
  "20240328", "Thursday", "MIL", "NYN", "d", "Rain",
  "20240328", "Thursday", "ANA", "BAL", "d", NA
) |>
  knitr::kable(caption = "2024 schedule (first 4 rows)")
```

### Day Games

The `day_night` column marks scheduled day ("d") and night ("n") games:

```{r, eval = FALSE}
schedule_2024 |>
  filter(day_night == "d") |>
  count(home_team, sort = TRUE) |>
  head(5)
```

```{r, echo = FALSE}
tribble(
  ~home_team, ~n,
  "CHN", 50L,
  "SDN", 40L,
  "OAK", 39L,
  "DET", 38L,
  "BAL", 37L
) |>
  knitr::kable(caption = "Teams with the most scheduled home day games in 2024")
```

### Games by Day of Week

```{r, eval = FALSE}
schedule_2024 |>
  count(day_of_week, sort = TRUE)
```

```{r, echo = FALSE}
tribble(
  ~day_of_week, ~n,
  "Saturday", 406L,
  "Sunday", 402L,
  "Friday", 391L,
  "Wednesday", 375L,
  "Tuesday", 372L,
  "Monday", 253L,
  "Thursday", 231L
) |>
  knitr::kable(caption = "Scheduled games by day of week in 2024")
```

### Postponed Games

The `postponed` column holds the reason when a game was postponed (35 games in 2024, mostly rain), and `makeup_date` says when it was rescheduled:

```{r, eval = FALSE}
schedule_2024 |>
  filter(!is.na(postponed)) |>
  select(date, visiting_team, home_team, postponed, makeup_date)
```

## Combined Analysis

### Join Game Logs with Rosters

```{r, eval = FALSE}
# Winning pitchers on the Yankees roster
yankees_wins <- gamelogs_2024 |>
  filter(home_team == "NYA" | visiting_team == "NYA") |>
  left_join(
    rosters_2024 |> select(player_id, last_name, first_name, team),
    by = c("winning_pitcher_id" = "player_id")
  ) |>
  filter(team == "NYA")
```

### Home vs Away Record

```{r, eval = FALSE}
gamelogs_2024 |>
  filter(home_team == "NYA" | visiting_team == "NYA") |>
  mutate(
    location = if_else(home_team == "NYA", "Home", "Away"),
    won = if_else(
      home_team == "NYA",
      home_score > visiting_score,
      visiting_score > home_score
    )
  ) |>
  group_by(location) |>
  summarize(games = n(), wins = sum(won), win_pct = mean(won))
```

## Performance Tips

- **Game logs are much faster than events.** For summary statistics (scores, decisions, attendance), `get_gamelogs()` downloads a fraction of the data that `get_events()` does.
- **Rosters reuse the event cache.** If you've already called `get_events()` for a year, `get_rosters()` for that year reads from the cached archive without downloading.
- **Download once, analyze many times.** All downloads are cached across R sessions; use `cache_status()` to see what you have locally.

## See Also

- `?get_gamelogs`, `?get_rosters`, `?get_schedules`
- The **Getting Started** article for the play-by-play event workflow
