---
title: "Getting Started with retrosheetshow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with retrosheetshow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo = FALSE, message = FALSE}
library(tibble)
library(dplyr)
```

> **RETROSHEET DATA NOTICE**
>
> The information used here was obtained free of charge from and is copyrighted by Retrosheet. Interested parties may contact Retrosheet at 20 Sunset Rd., Newark, DE 19711.

## Introduction

**retrosheetshow** provides a convenient and tidy interface for accessing [Retrosheet](https://www.retrosheet.org) baseball data in R. This vignette gets you started with the basic functionality.

Because every function downloads data from Retrosheet, the code chunks in this vignette are not evaluated when the package is built. The tables shown are real outputs captured from the 2024 season.

```{r setup, eval = FALSE}
library(retrosheetshow)
library(dplyr)
```

## Quick Start

### List Available Data

Use `list_events()` to discover what data is available:

```{r, eval = FALSE}
# List recent regular season files
list_events(year = 2020:2024)
```

```{r, echo = FALSE}
tribble(
  ~year, ~type, ~url,
  2024, "regular", "https://www.retrosheet.org/events/2024eve.zip",
  2023, "regular", "https://www.retrosheet.org/events/2023eve.zip",
  2022, "regular", "https://www.retrosheet.org/events/2022eve.zip",
  2021, "regular", "https://www.retrosheet.org/events/2021eve.zip",
  2020, "regular", "https://www.retrosheet.org/events/2020eve.zip"
) |>
  knitr::kable()
```

### Download Play-by-Play Events

`get_events()` downloads an event archive, extracts it, and parses every record. Downloads are cached, so only the first call for a given year touches the network. The postseason archives are small and make a quick first example:

```{r, eval = FALSE}
events <- get_events(year = 2024, type = "post")
```

The result has one row per record in the event files, with the raw fields in `f1`-`f6`. You'll usually hand it to one of the extractor functions below rather than work with it directly.

### Extract Game Information

`get_game_info()` returns one row per game with all its metadata:

```{r, eval = FALSE}
games <- get_game_info(events)
games |>
  select(game_id, date, visteam, hometeam, site, attendance)
```

```{r, echo = FALSE}
tribble(
  ~game_id, ~date, ~visteam, ~hometeam, ~site, ~attendance,
  "NYA202410140", "2024/10/14", "CLE", "NYA", "NYC21", "47264",
  "NYA202410150", "2024/10/15", "CLE", "NYA", "NYC21", "47054",
  "CLE202410170", "2024/10/17", "NYA", "CLE", "CLE08", "32531",
  "CLE202410180", "2024/10/18", "NYA", "CLE", "CLE08", "35263"
) |>
  knitr::kable(caption = "Game information (2024 ALCS, first 4 games)")
```

`get_plays()` returns one row per play, with the inning, batting team (0 = visiting, 1 = home), batter, count, pitch sequence, and the Retrosheet event string:

```{r, eval = FALSE}
plays <- get_plays(events)
plays |>
  select(game_id, inning, team, player_id, count, pitches, event)
```

```{r, echo = FALSE}
tribble(
  ~game_id, ~inning, ~team, ~player_id, ~count, ~pitches, ~event,
  "NYA202410140", 1L, 0L, "kwans001", "32", "BFFFBFFBX", "7/F7S",
  "NYA202410140", 1L, 0L, "fry-d001", "20", "BBX", "S7/G56",
  "NYA202410140", 1L, 0L, "ramij003", "21", "BBSX", "9/F89",
  "NYA202410140", 1L, 0L, "thoml002", "22", "BCSBB", "PB.1-2"
) |>
  knitr::kable(caption = "Play-by-play records (2024 ALCS game 1)")
```

The event strings (`7/F7S`, `S7/G56`, ...) are Retrosheet's play notation; see the **Retrosheet Format Reference** article for how to read them.

## Game Logs (Faster Alternative)

Game logs contain one row per game with 161 summary fields, and are much smaller than the play-by-play archives. Scores and counting statistics are parsed as integers:

```{r, eval = FALSE}
gamelogs <- get_gamelogs(year = 2024)
gamelogs |>
  select(date, visiting_team, home_team, visiting_score, home_score,
         winning_pitcher_name, attendance)
```

```{r, echo = FALSE}
tribble(
  ~date, ~visiting_team, ~home_team, ~visiting_score, ~home_score, ~winning_pitcher_name, ~attendance,
  "20240320", "LAN", "SDN", 5L, 2L, "Daniel Hudson", 15952L,
  "20240321", "SDN", "LAN", 15L, 11L, "Michael King", 15952L,
  "20240328", "COL", "ARI", 1L, 16L, "Zac Gallen", 48559L,
  "20240328", "WAS", "CIN", 2L, 8L, "Frankie Montas", 43799L
) |>
  knitr::kable(caption = "2024 game logs (first 4 games)")
```

```{r, eval = FALSE}
# Home field advantage in 2024
gamelogs |>
  summarize(
    total_games = n(),
    home_wins = sum(home_score > visiting_score),
    home_win_pct = mean(home_score > visiting_score)
  )
```

```{r, echo = FALSE}
tibble(total_games = 2429L, home_wins = 1267L, home_win_pct = 0.522) |>
  knitr::kable(caption = "Home field advantage in 2024")
```

## Team Rosters

```{r, eval = FALSE}
yankees <- get_rosters(year = 2024, team = "NYA")
yankees |>
  arrange(last_name)
```

```{r, echo = FALSE}
tribble(
  ~player_id, ~last_name, ~first_name, ~bats, ~throws, ~team, ~position,
  "andrc002", "Andrews", "Clayton", "L", "L", "NYA", "P",
  "beetc001", "Beeter", "Clayton", "R", "R", "NYA", "P",
  "bertj001", "Berti", "Jon", "R", "R", "NYA", "3B",
  "bickp001", "Bickford", "Phil", "R", "R", "NYA", "P"
) |>
  knitr::kable(caption = "2024 Yankees roster (first 4 players)")
```

## Game Schedules

```{r, eval = FALSE}
schedule <- get_schedules(year = 2024)
schedule |>
  select(date, game_number, day_of_week, visiting_team, home_team,
         day_night, postponed)
```

```{r, echo = FALSE}
tribble(
  ~date, ~game_number, ~day_of_week, ~visiting_team, ~home_team, ~day_night, ~postponed,
  "20240320", 0L, "Wednesday", "LAN", "SDN", "n", NA,
  "20240321", 0L, "Thursday", "SDN", "LAN", "n", NA,
  "20240328", 0L, "Thursday", "MIL", "NYN", "d", "Rain",
  "20240328", 0L, "Thursday", "ANA", "BAL", "d", NA
) |>
  knitr::kable(caption = "2024 schedule (first 4 rows)")
```

## Reference Data

```{r, eval = FALSE}
# Ballpark codes
parks <- get_park_ids()

# Team codes for a year
teams <- get_team_ids(2024)

# Player biographical database
players <- get_player_ids()
```

```{r, echo = FALSE}
tribble(
  ~park_id, ~name, ~city, ~state, ~start, ~league,
  "ATL03", "Suntrust Park", "Atlanta", "GA", "04/14/2017", "NL",
  "BOS07", "Fenway Park", "Boston", "MA", "04/20/1912", "AL",
  "CHI11", "Wrigley Field", "Chicago", "IL", "04/23/1914", "NL",
  "NYC21", "Yankee Stadium II", "New York", "NY", "04/16/2009", "AL"
) |>
  knitr::kable(caption = "Ballpark reference data (selected parks)")
```

## Caching

Downloads are cached under `tools::R_user_dir("retrosheetshow", "cache")`, so repeated calls for the same year read from disk instead of the network:

```{r, eval = FALSE}
# View cached files
cache_status()

# Remove cached files (by year/type, or everything)
clear_cache(year = 2024, type = "regular")
clear_cache()

# Disable caching for this session
use_cache(FALSE)
```

Regular-season event archives are the largest downloads (a few MB compressed per season); game logs and schedules are much smaller. If a download is interrupted, nothing is written to the cache, so simply retry.

## Complete Example

The 2024 Yankees, from game logs alone:

```{r, eval = FALSE}
library(retrosheetshow)
library(dplyr)

gamelogs <- get_gamelogs(year = 2024)

gamelogs |>
  filter(home_team == "NYA" | visiting_team == "NYA") |>
  mutate(
    yankees_score = if_else(home_team == "NYA", home_score, visiting_score),
    opponent_score = if_else(home_team == "NYA", visiting_score, home_score),
    won = yankees_score > opponent_score
  ) |>
  summarize(games = n(), wins = sum(won), win_pct = mean(won))
```

```{r, echo = FALSE}
tibble(games = 162L, wins = 94L, win_pct = 0.58) |>
  knitr::kable(caption = "2024 Yankees record")
```

## Next Steps

- See the **Game Logs, Rosters & Schedules** article for more examples
- Check the **Retrosheet Format Reference** to understand the event data structure
- Browse the **Reference** documentation for all available functions

## Attribution

Remember to cite Retrosheet when using this data:

> The information used here was obtained free of charge from and is copyrighted by Retrosheet. Interested parties may contact Retrosheet at 20 Sunset Rd., Newark, DE 19711.
